---
title: "ggpointdensity"
output:
  github_document
editor_options:
  chunk_output_type: inline
---
```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '', fig.width = 8, fig.height = 4, out.width = "100%", dpi=300
)
```

[![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/ggpointdensity)](https://cran.r-project.org/package=ggpointdensity) 
[![Downloads](https://cranlogs.r-pkg.org/badges/last-month/ggpointdensity?color=brightgreen)](https://cran.r-project.org/package=ggpointdensity)

Introduces `geom_pointdensity()`: A cross between a scatter plot and a 2D density plot.
```{r, include = FALSE}
library(tidyverse)
library(viridis)
library(ggpointdensity)
library(patchwork)
theme_set(theme_minimal())
```

```{r generate-toy-data, include = FALSE}
dat <- bind_rows(
  tibble(x = rnorm(7000, sd = 1),
         y = rnorm(7000, sd = 10),
         group = "foo"),
  tibble(x = rnorm(3000, mean = 1, sd = .5),
         y = rnorm(3000, mean = 7, sd = 5),
         group = "bar"))
```

```{r logo, echo =FALSE, fig.width=2.5, fig.height=2.5 , out.width="60%"}
dat %>%
  ggplot(aes(x = x, y = y)) +
  geom_pointdensity(size = .3) +
  scale_color_viridis() +
  labs(title="geom_pointdensity()") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none", aspect.ratio=1)
```
```{r, eval=FALSE, echo=FALSE}
dat %>%
  ggplot( aes( x = x, y = y, color = group)) +
  geom_point( size = .5)
```

## Installation
To install the package, type this command in R:
```{r, eval = FALSE}
install.packages("ggpointdensity")

# Alternatively, you can install the latest
# development version from GitHub:
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")
remotes::install_github("LKremer/ggpointdensity")
```

## Motivation
There are several ways to visualize data points on a 2D coordinate system:
If you have lots of data points on top of each other, `geom_point()` fails to
give you an estimate of how many points are overlapping.
`geom_density2d()` and `geom_bin2d()` solve this issue, but they make it impossible
to investigate individual outlier points, which may be of interest.

```{r, echo=FALSE, fig.width=10, fig.height=4}
dat %>%
  ggplot( aes( x = x, y = y)) +
  geom_point( size = .3) +
  labs(title="geom_point()") +

dat %>%
  ggplot( aes( x = x, y = y, fill = after_stat(level))) +
  stat_density_2d(geom = "polygon") +
  scale_fill_viridis() +
  labs(title="stat_density2d(geom='polygon')") +

dat %>%
  ggplot( aes( x = x, y = y)) +
  geom_bin2d() +
  scale_fill_viridis() +
  labs(title="geom_bin2d()") &

  theme(plot.title = element_text(hjust = 0.5), aspect.ratio = 1)
```

`geom_pointdensity()` aims to solve this problem by combining the best of both
worlds: individual points are colored by the number of neighboring points.
This allows you to see the overall distribution, as well as individual points.

```{r, echo = FALSE}
dat %>%
  ggplot(aes(x = x, y = y)) +
  geom_pointdensity(size = .3) +
  scale_color_viridis() +
  labs(title="geom_pointdensity()") +
  theme(plot.title = element_text(hjust = 0.5), aspect.ratio=1)
```

## Changelog
Added `method` argument and renamed the `n_neighbor` stat to `density`. The available options
are `method="auto"`,
`method="default"` and `method="kde2d"`. `default` is the regular n_neighbor calculation
as in the CRAN package. `kde2d` uses 2D kernel density estimation to estimate the point density
(credits to @slowkow).
This method is slower for few points, but faster for many (ca. >20k) points. By default,
`method="auto"` picks either `kde2d` or `default` depending on the number of points.

## Demo
Generate some toy data and visualize it with `geom_pointdensity()`:
```{r simple, include=TRUE, eval=TRUE, echo=TRUE}
<<generate-toy-data>>

ggplot(data = dat, mapping = aes( x = x, y = y)) +
  geom_pointdensity() +
  scale_color_viridis() +
  theme(aspect.ratio=1)
```


Each point is colored according to the number of neighboring points.
(Note: this here is the dev branch, where I decided to plot the density estimate
instead of n_neighbors now.)
The distance threshold to consider two points as neighbors (smoothing
bandwidth) can be adjusted with the `adjust` argument, where `adjust = 0.5`
means use half of the default bandwidth.
```{r adjusting-the-bandwidth, fig.width=4, fig.height=3, out.width="45%", fig.show="hold"}
ggplot(data = dat, mapping = aes(x = x, y = y)) +
  geom_pointdensity(size = .3, adjust = .1) +
  scale_color_viridis() +
  labs(title="adjust = 0.1") +
  theme(plot.title = element_text(hjust = 0.5), aspect.ratio=1)

ggplot(data = dat, mapping = aes(x = x, y = y)) +
  geom_pointdensity(size = .3, adjust = 4, aspect.ratio=1) +
  scale_color_viridis() +
  labs(title="adjust = 4") +
  theme(plot.title = element_text(hjust = 0.5), aspect.ratio=1)
```

Of course you can combine the geom with standard `ggplot2` features
such as facets...

```{r facets}
dat %>%
  ggplot( aes( x = x, y = y)) +
  geom_pointdensity(aes(color=after_stat(ndensity)), size = .25) +
  scale_color_viridis() +
  facet_wrap( ~ group) +
  labs(title="facet_wrap( ~ group)") +
  theme(plot.title = element_text(hjust = 0.5), aspect.ratio=1)
```

... or point shape and size:
```{r different-shapes}
dat_subset <- sample_frac(dat, .1)  # smaller data set
ggplot(data = dat_subset, mapping = aes(x = x, y = y)) +
  geom_pointdensity(size = 3, shape = 17) +
  scale_color_viridis() +
  labs(title="changing shape") +
  theme(plot.title = element_text(hjust = 0.5), aspect.ratio = 1)
```

Zooming into the axis works as well, keep in mind that `xlim()` and
`ylim()` change the density since they remove data points.
It may be better to use `coord_cartesian()` instead.

```{r zooming}
dat %>%
  ggplot(aes(x = x, y = y)) +
  geom_pointdensity(size = .5) +
  scale_color_viridis() +
  scale_x_continuous(limits = c(-1, 3)) +
  scale_y_continuous(limits = c(-5, 15)) +
  labs(title="using x- and ylim()") +

dat %>%
  ggplot(aes(x = x, y = y)) +
  geom_pointdensity(size = .5) +
  scale_color_viridis() +
  coord_cartesian(xlim = c(-1, 3), ylim = c(-5, 15)) +
  labs(title="using coord_cartesian()") &
  theme(aspect.ratio = 1, plot.title = element_text(hjust = 0.5))
```

```{r propotional-ink, eval=FALSE, echo = FALSE}
dat %>%
  ggplot(aes(x = x, y = y, size = after_stat(1/density), color = after_stat(density))) +
  geom_pointdensity(adjust = .2) +
  scale_color_viridis(option = "inferno", end = .9, direction = -1) +
  scale_size_area(max_size = 3) +
  theme(aspect.ratio = 1)
```

## Authors
Lukas PM Kremer ([@LPMKremer](https://twitter.com/LPMKremer/)) and Simon Anders ([@s_anders_m](https://twitter.com/s_anders_m/)), 2019
